---
title: "Colorectal Cancer in Wisconsin"
author: "Joule Voelz"
date: "`r Sys.Date()`"
output: html_document
---

# Load and join data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load required packages.

```{r, echo=TRUE, message=FALSE}
library(foreign)  
library(gdata)  
library(sp)
library(sf)
library(readr)
library(dplyr)
library(ggplot2)
library(SpatialEpi)
library(spdep)
library(INLA)
library(texreg)
library(stargazer)
library(xtable)
library(spdep)
```

Load Wisconsin shapefile downloaded from https://data-wi-dnr.opendata.arcgis.com/datasets/wi-dnr::county-boundaries-24k/about 

```{r, echo=TRUE,message=FALSE}
map <- st_read(
  "County_Boundaries_24K/County_Boundaries_24K.shp")
map <- st_as_sf(map)
```
Read in cleaned data.

```{r, echo=TRUE, message=FALSE}
covariates <- read_csv("data/data_cleaned/covariates.csv")
incidence <- read_csv("data/data_cleaned/incidence.csv")
mortality <- read_csv("data/data_cleaned/mortality.csv")
pop <- read_csv("data/data_cleaned/pop.csv")
land_area <- read_csv("data/data_cleaned/land_area.csv")
```
Join data to the map.

```{r}
map <- map %>% 
  merge(covariates,by.x=c("COUNTY_NAM"), by.y=c("County")) %>%
  merge(incidence,by.x=c("COUNTY_NAM"), by.y=c("County")) %>%
  merge(mortality,by.x=c("COUNTY_NAM"), by.y=c("County")) %>%
  merge(pop,by.x=c("COUNTY_NAM"), by.y=c("County")) %>%
  merge(land_area,by.x=c("COUNTY_NAM"), by.y=c("County"))
```


# Calculate expected cases and mortality

Calculate some statistics about statewide cases and mortality.

```{r}
wis_incidence <- incidence %>%
  filter(County == "Wisconsin")

wis_mortality <- mortality %>%
  filter(County == "Wisconsin")

wis_pop <- pop %>%
  filter(County == "Wisconsin")

wis_rate_all <- (wis_incidence["incidence_all"]/wis_pop["pop_all_adults"])[1,1]
wis_rate_under_65 <- (wis_incidence["incidence_under_65"]/wis_pop["pop_under_65"])[1,1]
wis_rate_65_over <- (wis_incidence["incidence_65_over"]/wis_pop["pop_65_over"])[1,1]

wis_mortality_rate_65_over <- (wis_mortality["mortality_65_over"]/wis_pop["pop_65_over"])[1,1]

wis_mortality_rate <- (wis_mortality["mortality_all"]/wis_pop["pop_all_adults"])[1,1]
```

Calculate expected cases and mortality.

```{r}
map$E_cases <- map$pop_under_65 * wis_rate_under_65 + map$pop_65_over * wis_rate_65_over

map$E_mortality_65_over <- map$pop_65_over * wis_mortality_rate_65_over

map$E_mortality_all <- map$pop_all_adults * wis_mortality_rate
```

Calculate SIR and SMR for each county.

```{r}
map$SIR <- map$incidence_all / map$E_cases
map$SMR_65_over <- map$mortality_65_over / map$E_mortality_65_over
map$SMR_all <- map$mortality_all / map$E_mortality_all
```

# Prepare and run models using INLA

Create neighborhood matrix and graph for INLA.

```{r}
nb <- poly2nb(map)
```

```{r}
nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")
```

Set random effects.

```{r}
map$re_u <- 1:nrow(map)
map$re_v <- 1:nrow(map)
```

Create formula for model.

```{r}
formula <- incidence_all ~ smokers_ever + poverty + endoscopy + f(re_u, model = "besag", graph = g, scale.model = TRUE) + f(re_v, model = "iid")
```

Create mask to exclude missing data points.

```{r}
mask <- !is.na(map$incidence_all)
```

Run INLA model.

```{r, echo=FALSE}
res <- inla(formula, family = "poisson", data = filter(map, !is.na(incidence_all)), E = E_cases, control.predictor = list(compute = TRUE))
```
```{r}
summary(res)
```
```{r}
xtable(res$summary.fixed[,c(1,2,3,5)], caption="Fitted coefficients", label="coefs",digits=3)
```

Plot coefficients.

```{r}
marginal <- inla.smarginal(res$marginals.fixed$endoscopy)
marginal <- data.frame(marginal)
ggplot(marginal, aes(x = x, y = y)) + geom_line() + labs(x = expression(beta[3]), y = "Density") +
  geom_vline(xintercept = 0, col = "blue") + theme_bw()
```

```{r}
marginal <- inla.smarginal(res$marginals.fixed$smokers_ever)
marginal <- data.frame(marginal)
ggplot(marginal, aes(x = x, y = y)) + geom_line() + labs(x = expression(beta[1]), y = "Density") +
  geom_vline(xintercept = 0, col = "blue") + theme_bw()
```
```{r}
marginal <- inla.smarginal(res$marginals.fixed$poverty)
marginal <- data.frame(marginal)
ggplot(marginal, aes(x = x, y = y)) + geom_line() + labs(x = expression(beta[2]), y = "Density") +
  geom_vline(xintercept = 0, col = "blue") + theme_bw()
```

Add fitted values to map.

```{r}
head(res$summary.fitted.values)
map$RR[mask] <- res$summary.fitted.values[, "mean"]
#map$LL[is.na(map$incidence_all)] <- res$summary.fitted.values[, "0.025quant"]
#map$UL[is.na(map$incidence_all)] <- res$summary.fitted.values[, "0.975quant"]
map$RR[!mask] <- NA
```

Run the same model for mortality.

```{r}
formula_mortality <- mortality_all ~ smokers_ever + poverty + endoscopy + f(re_u, model = "besag", graph = g, scale.model = TRUE) + f(re_v, model = "iid")
```
```{r}
mortality_mask <- !is.na(map$mortality_all)
```

```{r}
res_mortality <- inla(formula_mortality, family = "poisson", data = filter(map,mortality_mask), E = E_mortality_all, control.predictor = list(compute = TRUE))
```
```{r}
summary(res_mortality)
```
```{r}
xtable(res_mortality$summary.fixed[,c(1,2,3,5)], caption="Fitted coefficients (Mortality)", label="coefs_mortality",digits=3)
```

```{r}
head(res_mortality$summary.fitted.values)
map$RR_mortality[mortality_mask] <- res_mortality$summary.fitted.values[, "mean"]
#map$LL_mortality[mortality_mask] <- res_mortality$summary.fitted.values[, "0.025quant"]
#map$UL_mortality[mortality_mask] <- res_mortality$summary.fitted.values[, "0.975quant"]
map$RR_mortality[!mortality_mask] <- NA
```

# Generate maps

```{r}
ggplot(map) + 
  geom_sf(aes(fill = RR)) + 
  theme_bw() +  
  scale_fill_viridis_c(option = "C") +
  ggtitle("Relative Risk of Colorectal Cancer Incidence (Estimated)")

ggsave("figures/RR.pdf")
```

```{r, eval=FALSE}
pdf(file="figures/error_SIR.pdf")
plot(map$RR - map$SIR,
     ylab = "RR - SIR",
     main = "Error in Estimated Relative Risk (Incidence)")
abline(h=0, col="blue")
dev.off()
```
```{r, eval=FALSE}
pdf(file="figures/error_SMR.pdf")
plot(map$RR_mortality - map$SMR_all,
     ylab = "RR - SMR",
     main = "Error in Estimated Relative Risk (Mortality)")
abline(h=0, col="blue")
dev.off()
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = RR_mortality)) + 
  theme_bw() +  
  scale_fill_viridis_c(option = "C") +
  ggtitle("Relative Risk of Mortality from Colorectal Cancer (Estimated)") +
  labs(fill = "RR")

ggsave("figures/RR_mortality.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = SIR)) + 
  theme_bw() +  
  scale_fill_viridis_c(option = "C") +
  ggtitle("Standardized Incidence Ratios - Colorectal Cancer")

ggsave("figures/SIR.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = SMR_65_over)) + 
  theme_bw() +  
  scale_fill_viridis_c(option = "C") +
  ggtitle("Standardized Mortality Ratios - 65 and over")

#ggsave("figures/SIR.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = SMR_all)) + 
  theme_bw() +  
  scale_fill_viridis_c(option = "C") +
  ggtitle("Standardized Mortality Ratios") +
  labs(fill = "SMR")

ggsave("figures/SMR.pdf")
```
```{r}



```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = smokers_current)) + 
  theme_bw() +  
  scale_fill_viridis_c(option = "A") +
  ggtitle("Percent of Adults 18+ Who Are Currently Smokers")

ggsave("figures/current_smokers.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = smokers_ever)) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "A") +
  ggtitle("Percent of Adults 18+ Who Have Ever Been Smokers")

ggsave("figures/ever_smokers.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = poverty)) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "A") +
  ggtitle("Percent of Individuals Below Poverty Line")

ggsave("figures/poverty.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = endoscopy)) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "A") +
  ggtitle("Percent of Adults 50+ Ever Received Colorectal Endoscopy")

ggsave("figures/endoscopy.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = pop_65_over/pop_all_adults)) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "D") +
  ggtitle("Percentage of Population over 65") +
  labs(fill = "% over 65")

ggsave("figures/pop_over_65.pdf")
```
```{r}
ggplot(map) + 
  geom_sf(aes(fill = log(pop_all_adults))) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "D") +
  ggtitle("Log Population") +
  labs(fill = "Log Pop")

ggsave("figures/pop_log.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = log(pop_all_adults/land_area))) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "D") +
  ggtitle("Log Population Density (All Adults)") +
  labs(fill = "Log Pop Density")

ggsave("figures/pop_density.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = pop_all_adults/SHAPEAREA)) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "D") +
  ggtitle("Population Density") +
  labs(fill = "Pop Density")

#ggsave("figures/pop_log.pdf")
```

```{r}
ggplot(map) + 
  geom_sf(aes(fill = endoscopy)) + 
  theme_bw() + 
  scale_fill_viridis_c(option = "A") +
  ggtitle("Percent of Adults 50+ Ever Received Colorectal Endoscopy")

ggsave("figures/endoscopy.pdf")
```

# Calculate Morans

```{r}
rownames(map) <- map$COUNTY_NAM
nb <- poly2nb(filter(map,mask), queen = TRUE) # queen: sharing a point or a border
nbw <- spdep::nb2listw(nb, style = "W")
gmoran <- moran.test(filter(map,mask)$SIR, nbw)
gmoran
```

```{r}
pdf("figures/moran_sir.pdf")
moran.plot(filter(map,mask)$SIR, nbw, 
           labels = TRUE,
           xlab = "SIR",
           ylab = "Spatially Lagged SIR",
           main = "Moran's I - SIR",
           xlim=c(0.8, 1.6),
           ylim=c(0.8, 1.6))
dev.off()
```

```{r, eval=FALSE}
nb <- poly2nb(filter(map,mask), queen = TRUE) # queen: sharing a point or a border
nbw <- spdep::nb2listw(nb, style = "W")
gmoran <- moran.test(filter(map,mask)$RR, nbw)
gmoran
```

```{r, eval=FALSE}
pdf("figures/moran_rr.pdf")
moran.plot(filter(map,mask)$RR, nbw, 
           labels = TRUE,
           main = "Moran's I - Estimated RR (Incidence)",
           xlab = "RR",
           ylab = "Spatially Lagged RR",
           xlim=c(0.8, 1.6),
           ylim=c(0.8, 1.6))
dev.off()
```
